---
title: "Ames Housing — Exploratory Data Analysis"
author: "Christopher"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 9
    fig_height: 5.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,    # show code by default (collapsible via code_folding)
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

---

## 1  Introduction

This report explores the **Ames Housing dataset**, which contains information on
**2,930 residential home sales** in Ames, Iowa between 2006 and 2010. The dataset
was compiled by Dean De Cock as a modern alternative to the classic Boston Housing
dataset and is widely used for teaching regression and EDA.

Our goals are to:

- Understand the **distribution of sale prices**
- Explore how **living area** relates to price
- Compare **average prices across neighborhoods**
- Identify which numeric variables are most **correlated with sale price**

---

## 2  Setup & Data Loading

```{r packages}
# Install packages only if needed
required_packages <- c("tidyverse", "ggplot2", "corrplot", "scales", "AmesHousing")
new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if (length(new_packages) > 0) install.packages(new_packages)

library(tidyverse)
library(ggplot2)
library(corrplot)
library(scales)
library(AmesHousing)
```

```{r load-data}
# make_ames() returns a tidy tibble with clean column names
housing <- AmesHousing::make_ames()

cat("Dataset dimensions:", nrow(housing), "rows x", ncol(housing), "columns\n")
glimpse(housing)
```

---

## 3  Data Cleaning

```{r cleaning}
# --- Missing values ---
missing_counts <- colSums(is.na(housing))
missing_cols   <- missing_counts[missing_counts > 0]

if (length(missing_cols) == 0) {
  cat("No missing values detected.\n")
} else {
  cat("Columns with missing values:\n"); print(missing_cols)
}

# --- Impute & clean ---
housing_clean <- housing %>%
  mutate(across(where(is.numeric),   ~ ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  mutate(across(where(is.character),  ~ ifelse(is.na(.), "Unknown", .))) %>%
  mutate(across(where(is.factor),     ~ fct_na_value_to_level(., level = "Unknown")))

# --- Remove extreme price outliers (keep middle 99%) ---
price_bounds  <- quantile(housing_clean$Sale_Price, probs = c(0.005, 0.995))
housing_clean <- housing_clean %>%
  filter(Sale_Price >= price_bounds[1], Sale_Price <= price_bounds[2])

cat("Rows after cleaning:", nrow(housing_clean))
```

**Cleaning decisions made:**

| Step | Method | Reason |
|------|--------|--------|
| Numeric NAs | Replaced with column **median** | Median is robust to skewed distributions |
| Categorical NAs | Replaced with **"Unknown"** | Preserves the row without inventing a category |
| Price outliers | Removed bottom/top **0.5%** | Prevents extreme values from distorting plots |

---

## 4  Visualizations

### 4.1 Distribution of Sale Prices

```{r plot-price-dist}
# Consistent theme used across all plots
theme_ames <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(color = "grey40"),
      axis.title    = element_text(face = "bold"),
      plot.caption  = element_text(color = "grey50", size = 9)
    )
}

ggplot(housing_clean, aes(x = Sale_Price)) +
  geom_histogram(bins = 50, fill = "#4E79A7", color = "white", alpha = 0.85) +
  geom_vline(aes(xintercept = median(Sale_Price)),
             color = "#E15759", linetype = "dashed", linewidth = 1) +
  annotate("text",
           x = median(housing_clean$Sale_Price) * 1.05, y = Inf, vjust = 2,
           label = paste0("Median: $", comma(median(housing_clean$Sale_Price))),
           color = "#E15759", size = 4) +
  scale_x_continuous(labels = dollar_format()) +
  labs(
    title    = "Distribution of Home Sale Prices",
    subtitle = "Ames, Iowa — 2006 to 2010",
    x = "Sale Price (USD)", y = "Number of Homes",
    caption  = "Source: Ames Housing Dataset"
  ) +
  theme_ames()
```

**Insight:** Sale prices follow a **right-skewed distribution**: most homes sold
between \$100k–\$250k, but a long tail of luxury properties pulls the mean above
the median. The median sale price is approximately
**\$`r comma(median(housing_clean$Sale_Price))`**, which is a more representative
"typical" price than the mean for skewed data like this.

---

### 4.2 Sale Price vs. Living Area

```{r plot-scatter}
ggplot(housing_clean, aes(x = Gr_Liv_Area, y = Sale_Price)) +
  geom_point(alpha = 0.35, color = "#4E79A7", size = 1.5) +
  geom_smooth(method = "lm", color = "#E15759", se = TRUE, linewidth = 1.2) +
  scale_y_continuous(labels = dollar_format()) +
  scale_x_continuous(labels = comma_format()) +
  labs(
    title    = "Sale Price vs. Above-Ground Living Area",
    subtitle = "Each point = one home; red line = linear trend with 95% confidence band",
    x = "Above-Ground Living Area (sq ft)", y = "Sale Price (USD)",
    caption  = "Source: Ames Housing Dataset"
  ) +
  theme_ames()
```

**Insight:** There is a **clear positive linear relationship** between living area
and sale price: larger homes generally sell for more. The linear trend fits well
for most of the data, though a handful of large homes sold for surprisingly low
prices (possible distressed sales or unusual properties). The spread of prices
also widens as living area increases, a pattern called
**heteroscedasticity**, which is common in real estate data.

---

### 4.3 Average Sale Price by Neighborhood

```{r plot-neighborhood, fig.height=8}
neighborhood_avg <- housing_clean %>%
  group_by(Neighborhood) %>%
  summarise(avg_price = mean(Sale_Price), .groups = "drop") %>%
  arrange(desc(avg_price)) %>%
  mutate(Neighborhood = fct_reorder(Neighborhood, avg_price))

ggplot(neighborhood_avg, aes(x = Neighborhood, y = avg_price, fill = avg_price)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient(low = "#AEC6CF", high = "#1F4E79") +
  scale_y_continuous(labels = dollar_format()) +
  coord_flip() +
  labs(
    title    = "Average Sale Price by Neighborhood",
    subtitle = "Ranked from highest to lowest average price",
    x = NULL, y = "Average Sale Price (USD)",
    caption  = "Source: Ames Housing Dataset"
  ) +
  theme_ames()
```

**Insight:** Neighborhood is one of the strongest predictors of home price.
**`r neighborhood_avg$Neighborhood[1]`** commands the highest
average prices, while more affordable neighborhoods cluster at the bottom. The
price gap between the most and least expensive neighborhoods is over
**\$`r comma(max(neighborhood_avg$avg_price) - min(neighborhood_avg$avg_price))`**,
highlighting how much location matters in real estate.

---

### 4.4 Correlation Heatmap

```{r plot-heatmap, fig.width=8, fig.height=7}
key_numeric_vars <- c(
  "Sale_Price", "Gr_Liv_Area", "Total_Bsmt_SF", "Garage_Area",
  "Year_Built", "Year_Remod_Add", "Lot_Area", "Overall_Qual",
  "Overall_Cond", "TotRms_AbvGrd", "Bedroom_AbvGr", "Full_Bath"
)

cor_matrix <- housing_clean %>%
  select(all_of(key_numeric_vars)) %>%
  # Some columns (e.g. Overall_Qual, Overall_Cond) are ordered factors —
  # convert everything to numeric so cor() works
  mutate(across(everything(), ~ as.numeric(.))) %>%
  cor(use = "complete.obs")

corrplot(
  cor_matrix,
  method      = "color",
  type        = "upper",
  tl.cex      = 0.85,
  tl.col      = "black",
  addCoef.col = "black",
  number.cex  = 0.65,
  col         = colorRampPalette(c("#4E79A7", "white", "#E15759"))(200),
  title       = "Correlation Heatmap — Key Numeric Variables",
  mar         = c(0, 0, 2, 0)
)
```

**Insight:** The heatmap reveals several strong relationships with `Sale_Price`:

- **Overall_Qual** (overall material quality) has the **highest correlation**
  (~0.80), confirming that build quality is the single strongest numeric predictor.
- **Gr_Liv_Area**, **Total_Bsmt_SF**, and **Garage_Area** are all positively
  correlated: more space means more value.
- **Year_Built** and **Year_Remod_Add** show moderate positive correlations,
  meaning newer or recently renovated homes sell for more.
- **Overall_Cond** (overall condition) has a surprisingly **weak correlation**
  with price, unlike quality — suggesting buyers care more about the *grade* of
  materials than the *state of repair*.

---

## 5  Simple Linear Regression Model

Now that the EDA has identified the strongest predictors, we build a simple
multiple linear regression model to predict `Sale_Price` using three variables:

- `Gr_Liv_Area` — above-ground living area (sq ft)
- `Overall_Qual` — overall material & finish quality (ordered factor → numeric)
- `Garage_Cars` — garage capacity (number of cars)

```{r regression-prep}
# Convert Overall_Qual (ordered factor) to numeric for use in lm()
# The numeric levels preserve the quality ordering (1 = worst, 10 = best)
model_data <- housing_clean %>%
  mutate(Overall_Qual_num = as.numeric(Overall_Qual)) %>%
  select(Sale_Price, Gr_Liv_Area, Overall_Qual_num, Garage_Cars)
```

### 5.1 Model Summary

```{r regression-model}
# Fit the linear regression model
# lm(outcome ~ predictor1 + predictor2 + ..., data = your_data)
model <- lm(Sale_Price ~ Gr_Liv_Area + Overall_Qual_num + Garage_Cars,
            data = model_data)

summary(model)
```

### 5.2 Model Performance (RMSE)

```{r regression-rmse}
# RMSE = Root Mean Squared Error
# Measures the average prediction error in the same units as Sale_Price (dollars)
# Lower RMSE = better predictions

predictions <- predict(model, newdata = model_data)
residuals   <- model_data$Sale_Price - predictions
rmse        <- sqrt(mean(residuals^2))

cat("RMSE: $", comma(round(rmse, 0)), "\n", sep = "")
cat("R-squared:", round(summary(model)$r.squared, 3), "\n")
cat("Adjusted R-squared:", round(summary(model)$adj.r.squared, 3), "\n")
```

### 5.3 Predicted vs. Actual Sale Prices

```{r regression-plot}
# Build a data frame of actual vs. predicted values for plotting
pred_df <- tibble(
  actual    = model_data$Sale_Price,
  predicted = predictions
)

ggplot(pred_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.3, color = "#4E79A7", size = 1.5) +
  # Perfect predictions would fall on this diagonal line
  geom_abline(slope = 1, intercept = 0,
              color = "#E15759", linetype = "dashed", linewidth = 1) +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title    = "Predicted vs. Actual Sale Prices",
    subtitle = "Points close to the red line = accurate predictions",
    x        = "Actual Sale Price (USD)",
    y        = "Predicted Sale Price (USD)",
    caption  = "Model: Sale_Price ~ Gr_Liv_Area + Overall_Qual + Garage_Cars"
  ) +
  theme_ames()
```

**Interpretation:** The model explains approximately
**`r round(summary(model)$r.squared * 100, 1)`% of the variation** in sale prices
(R² = `r round(summary(model)$r.squared, 3)`), which is solid for just three
predictors. Each additional square foot of living area adds roughly
**\$`r comma(round(coef(model)["Gr_Liv_Area"], 0))`** to the predicted price,
each one-step increase in overall quality adds around
**\$`r comma(round(coef(model)["Overall_Qual_num"], 0))`**, and each additional
garage car space adds approximately
**\$`r comma(round(coef(model)["Garage_Cars"], 0))`**. The predicted vs. actual
plot shows points clustering tightly around the diagonal, good alignment in the
\$100k–\$300k range where most homes sit. The model tends to **underpredict**
for high-end homes (top-right scatter above the line), which is expected: luxury
properties have premium features (custom finishes, location) that a 3-variable
model can't fully capture. An RMSE of roughly \$`r comma(round(rmse/1000, 0))`k
means typical predictions are off by that amount, reasonable as a baseline, but
there's clear room to improve by adding neighborhood, year built, or basement
area as additional predictors.

---

## 6  Summary

| Question | Finding |
|----------|---------|
| What does the price distribution look like? | Right-skewed; median ~\$163k, mean pulled higher by luxury homes |
| Does bigger = pricier? | Yes, strong positive linear trend with living area |
| Which neighborhood is most expensive? | `r neighborhood_avg$Neighborhood[1]` leads; 28 neighborhoods show wide price variation |
| What drives sale price most? | Overall quality, living area, basement size, and garage area |

**Next steps for deeper analysis:**

1. Investigate the **outlier homes** (large area, low price) in the scatter plot.
2. Explore **time trends**, did prices change between 2006 and 2010?
3. Add more predictors (neighborhood, year built, basement area) to improve the model's RMSE.

---

*Report generated with R `r R.version$major`.`r R.version$minor` and
[`AmesHousing`](https://cran.r-project.org/package=AmesHousing) package.*
